---
title: "Practical Machine Learning - Quiz 3"
author: "Sanjay-Srivastava"
date: "October 13, 2015"
output: html_document
---

# Question 1
---------------------------------------------------------------------------------------------
Load the cell segmentation data from the AppliedPredictiveModeling package using the commands:

* library(AppliedPredictiveModeling)
* data(segmentationOriginal)
* library(caret)

1. Subset the data to a training set and testing set based on the Case variable in the data set. 
2. Set the seed to 125 and fit a CART model with the rpart method using all predictor variables and default caret settings. 
3. In the final model what would be the final model prediction for cases with the following variable values:
a. TotalIntench2 = 23,000; FiberWidthCh1 = 10; PerimStatusCh1=2 
b. TotalIntench2 = 50,000; FiberWidthCh1 = 10;VarIntenCh4 = 100 
c. TotalIntench2 = 57,000; FiberWidthCh1 = 8;VarIntenCh4 = 100 
d. FiberWidthCh1 = 8;VarIntenCh4 = 100; PerimStatusCh1=2 

---------------------------------------------------------------------------------------------

```{r PracticalMachineLearing_Quiz3_Question1}

library(AppliedPredictiveModeling)
data(segmentationOriginal)
library(caret)
library(rattle)
library(rpart)
## subset the data into test and training sets using the values in variable: Case
unique(segmentationOriginal$Case)
training = segmentationOriginal[segmentationOriginal$Case == "Train", ]
testing = segmentationOriginal[segmentationOriginal$Cell == "Test", ]

set.seed(125)
modfit = train(Class ~ ., data = training, method = "rpart")
modfit$finalModel
fancyRpartPlot(modfit$finalModel)

```

Note the following:

* The decision tree is constructed using only two variables: TotalIntenCh2 & FiberWdithCh1, No other variables are required by the model.
* When TotalIntenCh2 = 23000 (less than 45000), the value is PS and we don't need the other.
* TotalIntenCh2 = 50000 (> 45000), and FiberWidthCh1 = 10 ( > 9.7), the outcome is WS 
* Similarly, when TotalIntenCh2 = 57000 and FiberWidthCh1 = 8, the outcome is PS
* In the last case, neither of the two parameters are given, so the model can't predict an outcome.

# Question 3
---------------------------------------------------------------------------------------------

Load the olive oil data using the commands:

* library(pgmm)
* data(olive)
* olive = olive[,-1]

(NOTE: If you have trouble installing the pgmm package, you can download the olive dataset here: olive_data.zip. After unzipping the archive, you can load the file using the load() function in R.)
These data contain information on 572 different Italian olive oils from multiple regions in Italy. Fit a classification tree where Area is the outcome variable. Then predict the value of area for the following data frame using the tree command with all defaults

newdata = as.data.frame(t(colMeans(olive)))
 
What is the resulting prediction? Is the resulting prediction strange? Why or why not?

```{r PracticalMachineLearning_Quiz3_Question3}
   library(pgmm)
   data(olive)
   unique(olive$Area)
   training = olive[, -1] ## remove the Region column from the dataset
   testing = as.data.frame(t(colMeans(training)))
   modfit = train(data = training, Area ~ ., method = "rpart")
   predictions = predict(modfit, newdata = testing)
   print (paste("Predicted Area: ", round(predictions,2)))

```

Note:
Even though the variable Area is a categorical variable, the prediction is in decimals, because we did not convert Area into a factor variable. 

# Question 4
---------------------------------------------------------------------------------------------
Load the South Africa Heart Disease Data and create training and test sets with the following code:

* rm(list=ls())
* library(ElemStatLearn)
* data(SAheart)
* set.seed(8484)
* train = sample(1:dim(SAheart)[1],size=dim(SAheart)[1]/2,replace=F)
* trainSA = SAheart[train,]
* testSA = SAheart[-train,]

Then set the seed to 13234 and fit a logistic regression model (method="glm", be sure to specify family="binomial") with Coronary Heart Disease (chd) as the outcome and age at onset, current alcohol consumption, obesity levels, cumulative tabacco, type-A behavior, and low density lipoprotein cholesterol as predictors. Calculate the misclassification rate for your model using this function and a prediction on the "response" scale:

Define the mis-classification function as follows:
missClass = function(values,prediction){sum(((prediction > 0.5)*1) != values)/length(values)}

What is the misclassification rate on the training set? What is the misclassification rate on the test set?


```{r PracticalMachineLearning_Quiz3_Question4}
   
   ## the first command rm(list = ls()) is not required to be executed here. ls() returns
   ## the objects in the current environment. rm() just goes and flushes those objects from
   ## the memory
   
    library(ElemStatLearn)
    data(SAheart)
    set.seed(8484)
    train = sample(1:dim(SAheart)[1],size=dim(SAheart)[1]/2,replace=F)
    trainSA = SAheart[train,]
    testSA = SAheart[-train,]
    
    set.seed(13234)
    cols = c("chd", "age", "alcohol", "tobacco", "obesity", "typea", "ldl")
    trainSA = trainSA[, cols]
    ##trainSA$chd = as.factor(trainSA$chd)
    
    testSA = testSA[, cols]
    ##testSA$chd = as.factor(testSA$chd)
    
    ##define the miss-classification function:
    missClass = function(values,prediction){sum(((prediction > 0.5)*1) != values)/length(values)}
    
    modfit = train(data = trainSA, factor(chd) ~ ., method = 'glm', family = "binomial")
    predictions_train = predict(modfit, trainSA)
    predictions_test  = predict(modfit, testSA)
    
    missed_train = missClass(predictions_train, trainSA$chd)
    missed_test  = missClass(predictions_test, testSA$chd)
    
    print(paste(
                "Training Data Missclassification Rate: ", round(missed_train,2),
                "Test Data Missclassification Rate: ", round(missed_test,2)
                )
          )
    
```

# Question 5
---------------------------------------------------------------------------------------------

Load the vowel.train and vowel.test data sets:

* rm(list=ls())
* library(ElemStatLearn)
* data(vowel.train)
* data(vowel.test)
Set the variable y to be a factor variable in both the training and test set. Then set the seed to 33833. Fit a random forest predictor relating the factor variable y to the remaining variables. Read about variable importance in random forests here: 
[Random Forests](http://www.stat.berkeley.edu/~breiman/RandomForests/cc_home.htm#ooberr) 
The caret package uses by defualt the Gini importance. Calculate the variable importance using the varImp function in the caret package. What is the order of variable importance?



```{r PracticalMachineLearning_Quiz3_Question5}
    library(ElemStatLearn)
    library(randomForest)
    
    training = vowel.train
    testing = vowel.test
    training$y = as.factor(training$y)
    testing$y = as.factor(testing$y)
    
    set.seed(33833)
    modfit = randomForest(y ~ ., data = training, importance = FALSE)
    order(varImp(modfit), decreasing = TRUE)
    varImpPlot(modfit, sort = TRUE)
```   
